---
title: "Can Money Buy Higher Life Expectancy?"
author: "Group 14: Morgan Barney, Ana Porro, and Sarah Albert"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: kable
---

<!--
Comments in HTML are like this! 
-->

```{r, include=FALSE}
#Load Packages
library(ggplot2)
library(tidyverse)
library(leaflet)
library(sf)
library(USAboundaries)
library(maps)
library(dplyr)
library(plyr)
library(readr)
library(knitr)
library(htmltools)
library(htmlwidgets)
library(RColorBrewer)
library(magrittr)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Data Wrangling

#Load in shape file
cz_1990 <- "cz1990_shapefile" %>%
  read_sf()

#load in data sets
health_ineq_online_table_11 <- read_csv("health_ineq_online_table_11.csv")

health_ineq_online_table_12 <- read_csv("health_ineq_online_table_12.csv")

#select important variables from health_ineq_online_table_11
#this is so the full join goes nicely
health_ineq_online_table_11_v2 <- select(health_ineq_online_table_11, cty, statename, le_raceadj_q1_F, le_agg_q1_F, 
                                         le_raceadj_q2_F, le_agg_q2_F, le_raceadj_q3_F, le_agg_q3_F, 
                                         le_raceadj_q4_F, le_agg_q4_F, le_raceadj_q1_M, le_agg_q1_M, 
                                         le_raceadj_q2_M, le_agg_q2_M, le_raceadj_q3_M, le_agg_q3_M, 
                                         le_raceadj_q4_M, le_agg_q4_M)

#combine health_ineq_online_table_11 and health_ineq_online_table_12 by county (cty) and view
income_and_health_by_cty <- full_join(health_ineq_online_table_11_v2, health_ineq_online_table_12, by = "cty")

#combine income_and_health_by_cty and shapefile by communting zone (cz)
income_and_health_by_cz <- full_join(income_and_health_by_cty, cz_1990, by = "cz")

#filtering for NE States
income_and_health_by_cz_NE <- income_and_health_by_cz %>% filter(statename.x == "Massachusetts" | 
                                                                   statename.x == "Maine" |
                                                                   statename.x == "New Hampshire" |
                                                                   statename.x == "Vermont" |
                                                                   statename.x == "Rhode Island")
#View(income_and_health_by_cz_NE)
```


In the current political atmosphere, a hot topic has been universal healthcare access.  However, too often the discussion leaves out the substantial role that income plays in shaping one's health outcomes and associated behaviors.  This analysis aims to investigate the role of income inequality in such outcomes and health behaviors to further understand what health problems plague the poor and why. By doing this, we hope to create the most impactful policy framework to improve the health outcomes and  behaviors of the lowest income people.  The visualization below shows a preliminary overview of such health outcomes in New England. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height= 10/2, fig.width= 16/2 }
#First Graph: Life Expectancy by Income Quartile and Gender


#create new column coded with income quartiles for life expectency by income quartile
income_and_mortlity_rates_NE <- income_and_health_by_cz_NE %>% 
  select(cty, statename.x, le_agg_q1_F, 
         le_agg_q2_F, le_agg_q3_F, le_agg_q4_F, 
         le_agg_q1_M, le_agg_q2_M, le_agg_q3_M, le_agg_q4_M,
         geometry, rel_tot, 
         scap_ski90pcm, cs00_seg_inc_aff75, poor_share) %>%
  #gathering
  gather(key = income_quartile, value = life_expectancy, le_agg_q1_F,
         le_agg_q2_F, le_agg_q3_F, le_agg_q4_F, le_agg_q1_M, le_agg_q2_M,
         le_agg_q3_M, le_agg_q4_M) 

#code as factor to order facets
income_and_mortlity_rates_NE$income_quartile_GG = factor(income_and_mortlity_rates_NE$income_quartile, levels=c('le_agg_q1_F','le_agg_q2_F','le_agg_q3_F','le_agg_q4_F', 'le_agg_q1_M', 'le_agg_q2_M','le_agg_q3_M', 'le_agg_q4_M'))
#lables for facets
levels=c('le_agg_q1_F','le_agg_q2_F','le_agg_q3_F','le_agg_q4_F', 'le_agg_q1_M', 'le_agg_q2_M','le_agg_q3_M', 'le_agg_q4_M')

income_quartile_GG2 <- c(le_agg_q1_F = "Quartile 1 Females", le_agg_q2_F = "Quartile 2 Females", le_agg_q3_F = "Quartile 3 Females", le_agg_q4_F = "Quartile 4 Females", 
                         le_agg_q1_M = "Quartile 1 Males", le_agg_q2_M = "Quartile 2 Males", le_agg_q3_M = "Quartile 3 Males", le_agg_q4_M = "Quartile 4 Males")


#recode variable as numeric to assign color value
income_and_mortlity_rates_NE <- income_and_mortlity_rates_NE %>%
mutate(life_expectancy = as.numeric(life_expectancy))

#create new variable to assign color groups to
income_and_mortlity_rates_NE$life_expectancy_GROUPS <- cut(income_and_mortlity_rates_NE$life_expectancy, breaks = c( 70,73, 76, 79, 82, 85, 88, 91, Inf),
                            labels = c("70-72",
                                    "73-75",
                                    "76-78",
                                    "79-81",
                                    "82-84",
                                    "85-87",
                                    "88-90",
                                    "91 and Above"
                             ))

#plot life expectancy by income quartile and gender
ggplot() + geom_sf(data = income_and_mortlity_rates_NE, aes(fill = income_and_mortlity_rates_NE$life_expectancy_GROUPS)) + 
  facet_wrap(.~income_quartile_GG, labeller = labeller(income_quartile_GG=income_quartile_GG2), nrow = 2) + labs(title= "Life Expectancy per Income Quartile by Gender in New England", subtitle = "What role do income and gender play in life expectancy in the Northeast?", fill="Life\nExpectancy") + 
  coord_sf(xlim = c(-67, -74), ylim = c(41, 47.4)) +
 # ggtitle("Life Expectancy per Income Quartile by Gender in New England", subtitle = "What role do income and gender play in life expectancy in the Northeast?")+
  #change background to white
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "white"),
        #changes facet background and outline
        strip.background = element_rect(colour="white", fill="white"),
        #centers and bolds title and subtitle
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle =element_text(hjust = 0.5),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
     legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold")) +
      scale_fill_brewer(palette="YlOrRd") + 
  theme(legend.key.size = unit(4, "mm"))
     
#legend.key.size = unit(1, "strwidth") 

  #legend.key.size = 
```



This visualization displays the life expectancy of New Englanders colored by commuting zone in 2000. This graph is separated by both sex and income quartile where the lowest quartile is represented by 1, and the highest by 4. 

There are several takeaways from this analysis. Firstly, in New England, it appears that life expectancy per income quartile is not equal across gender. This visualization shows that on average, women have a higher life expectancy than men in all income quartiles. Additionally, life expectancy increases by income quartile.  Therefore, perhaps unsurprisingly, it appears that life expectancy may be a byproduct of income inequality. To further investigate the reasons for this trend, we investigated the health behaviors that we thought could be a consequence of income inequality. To read more about our results, click on the [more info tab!](https://aporro95.github.io/webpage/more_info.html) 
	 
The data used in this analysis was taken from the paper “The Association between Income and Life Expectancy in the United States, 2001-2014” by Raj Chetty et al.^[Chetty, Raj, et al. “The Association between Income and Life Expectancy in the United States, 2001 - 2014.” The Equality of Opportunity Project, Opportunity Insights, 2016, www.equality-of-opportunity.org/data/index.html#health.]  The authors constructed income quartiles using income data for the U.S. population through de-identified tax records between 1999-2014. Mortality data was obtained from Social Security Administration's death records.


