---
title: "The Effects of Smoking, Obesity, and Exercise on Life Expectancy"
author: "Morgan Barney, Sarah Albert, and Ana Porro"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable
---

<!--
Comments in HTML are like this! 
-->

```{r, include=FALSE, warning=FALSE, message=FALSE}
#Load Packages
library(ggplot2)
library(tidyverse)
library(leaflet)
library(sf)
library(USAboundaries)
library(maps)
library(dplyr)
library(plyr)
library(readr)
library(knitr)
library(htmltools)
library(htmlwidgets)
library(RColorBrewer)
library(magrittr)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Data Wrangling

#Load in shape file
cz_1990 <- "cz1990_shapefile" %>%
  read_sf()

#load in data sets
health_ineq_online_table_11 <- read_csv("health_ineq_online_table_11.csv")

health_ineq_online_table_12 <- read_csv("health_ineq_online_table_12.csv")

#select important variables from health_ineq_online_table_11
#this is so the full join goes nicely
health_ineq_online_table_11_v2 <- select(health_ineq_online_table_11, cty, statename, le_raceadj_q1_F, le_agg_q1_F, 
                                         le_raceadj_q2_F, le_agg_q2_F, le_raceadj_q3_F, le_agg_q3_F, 
                                         le_raceadj_q4_F, le_agg_q4_F, le_raceadj_q1_M, le_agg_q1_M, 
                                         le_raceadj_q2_M, le_agg_q2_M, le_raceadj_q3_M, le_agg_q3_M, 
                                         le_raceadj_q4_M, le_agg_q4_M)

#combine health_ineq_online_table_11 and health_ineq_online_table_12 by county (cty) and view
income_and_health_by_cty <- full_join(health_ineq_online_table_11_v2, health_ineq_online_table_12, by = "cty")

#combine income_and_health_by_cty and shapefile by communting zone (cz)
income_and_health_by_cz <- full_join(income_and_health_by_cty, cz_1990, by = "cz")

#filtering for NE States
income_and_health_by_cz_NE <- income_and_health_by_cz %>% filter(statename.x == "Massachusetts" | 
                                                                   statename.x == "Maine" |
                                                                   statename.x == "New Hampshire" |
                                                                   statename.x == "Vermont" |
                                                                   statename.x == "Rhode Island")
#View(income_and_health_by_cz_NE)
```




#Introduction

To further investigate the possible factors which may influence low-income peoples’ relatively depressed life expectancies, we used survey data from the Behavioral Risk Factor Surveillance System (BRFSS) during the year 2000, found in Chetty et. al’s data set.  The BRFSS is the nation's premier provider of health-related telephone surveys which collect state data about U.S. residents regarding their health-related behaviors.^[“Behavioral Risk Factor Surveillance System.” Centers for Disease Control and Prevention, National Center for Chronic Disease Prevention and Health Promotion , Division of Population Health, 16 May 2014, www.cdc.gov/brfss/index.html"]  The three health indicators we considered by income quartile were tobacco usage, obesity rates, and exercise rates.








#Tobacco
	
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height= 10, fig.width= 14}
#Graph 4: Smokers by Income Quartile

income_and_smoking_rates_NE <- income_and_health_by_cz_NE %>% 
  select(cty, statename.x, geometry, le_agg_q1_F, 
         cur_smoke_q1, cur_smoke_q2, cur_smoke_q3, cur_smoke_q4) %>%
  #gathering
  gather(key = income_quartile, value = perc_smokers,
         cur_smoke_q1, cur_smoke_q2, cur_smoke_q3, cur_smoke_q4)

#labels for smoking facets
smoke_quartile_labels <- c(cur_smoke_q1 = "Quartile 1", cur_smoke_q2 = "Quartile 2", cur_smoke_q3 = "Quartile 3", cur_smoke_q4 = "Quartile 4") 

#graph smoking rates by income

ggplot() + geom_sf(data = income_and_smoking_rates_NE, aes(fill = perc_smokers))+ 
  facet_wrap(~income_quartile, labeller = labeller(income_quartile = smoke_quartile_labels), nrow = 1) +
  coord_sf(xlim = c(-67, -74), ylim = c(41, 47.4)) +
  #changes color palette
  #scale_fill_brewer(palette="YlGn") +
  #change background to white
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "white"),
        #changes facet background and outline
        strip.background = element_rect(colour="white", fill="white"),
        #centers and bolds title and subtitle
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle =element_text(hjust = 0.5),
        #gets rid of x and y axis labels
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_text(face = "bold"),
        legend.direction = "horizontal",
       legend.position = "bottom") +
  #labels
  labs(title= "Smoking Rates Per Income Quartile in New England", fill = "Share of\nSmokers", subtitle ="How does income influence smoking?") +
   scale_fill_distiller(palette = "Reds", direction = 1)
       
  
  
  #scale_fill_gradientn(colours = heat.colors(4))
  
  

#\\ Smoking is inversely related to income, as visualized in the graph, there is a remarkable decrease \\ in #the percentage of smokers in New England as income steadily increases.")
```

Cigarettes pose a great risk of death and disease, specifically lung cancer and have been manufactured to be more addictive over the years. The smoking rate used to be higher among more educated people pre-WWII- before the health risks of smoking were widely known. After the first surgeon general’s report on smoking was issued in 1964, more educated individuals stopped smoking as much and the tobacco industry began targeting low-income individuals (Truth Initiative.) Tobacco retailers marketed their tobacco products to low-income individuals by attaching coupons to food stamps and giving out free cigarettes in housing projects.  The tobacco industry fights to keep prices low so that low-income people can continue to buy cigarettes by spending billions of dollars each year to fight cigarette tax and price increases (Truth Initiative.) Now, smoking is linked with lower levels of education and 72% of smokers are from low-income communities and more than a quarter of people living below the poverty level and three-quarters of homeless people use tobacco (Public HEalth Law Center.) A 2016 report on the economics of tobacco published by the National Cancer Institute and World Health Organization found that a significant amount of health disparities between the rich and poor can be attributed to the differences in tobacco habits between income groups. Compounding this effect is the fact that low-income communities have less access to resources to help them quit smoking, and are exposed to less education on the health hazards of smoking. 

https://truthinitiative.org/news/why-are-72-percent-smokers-lower-income-communities  
https://truthinitiative.org/news/smoking-and-low-income-communities 
https://publichealthlawcenter.org/sites/default/files/resources/tclc-homeless-tobacco-FAQ-2016.pdf 


#Obesity

The US is one of the wealthiest countries in the world yet it has one of the highest obesity rates. One-third of the population is obese while another third is overweight (American Diabetes Association.) Hunger and lack of availability of healthy foods are two contributing factors to the higher level of obesity among people living in poverty. Poverty-dense urban areas are oftentimes referenced as “food deserts” because of the lack of availability and affordability of fresh and good quality food. However, hunger and poor availability of healthy food is only one reason why obesity tends to track with poverty. 
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height= 10, fig.width= 14}
#Graph: Obesity by Income Quartile

#obesity rates by income
income_and_obesity_rates_NE <- income_and_health_by_cz_NE %>% 
  select(cty, statename.x, geometry, le_agg_q1_F, 
         bmi_obese_q1, bmi_obese_q2, bmi_obese_q3, bmi_obese_q4) %>%
  #gathering
  gather(key = income_quartile, value = perc_obese,
         bmi_obese_q1, bmi_obese_q2, bmi_obese_q3, bmi_obese_q4)

#code as factor to order facets
income_and_obesity_rates_NE$income_quartile_GG = factor(income_and_obesity_rates_NE$income_quartile, levels=c('bmi_obese_q1','bmi_obese_q2','bmi_obese_q3','bmi_obese_q4'))
                                                        #lables for facets
                                                        levels=c('bmi_obese_q1','bmi_obese_q2','bmi_obese_q3','bmi_obese_q4') 
                                                        
income_and_obesity_rates_NE2 <- c(bmi_obese_q1 = "Quartile 1", bmi_obese_q2 = "Quartile 2", bmi_obese_q3 = "Quartile 3", bmi_obese_q4 = "Quartile 4")

#factor bmi_obesity for color scale

#graph obsesity rates
ggplot() + geom_sf(data = income_and_obesity_rates_NE, aes(fill = perc_obese))+ 
            facet_wrap(~income_quartile, labeller = labeller(income_quartile = income_and_obesity_rates_NE2), nrow = 1) +
            labs(title = "Obesity Per Income Quartile in New England",  fill = "Percent Obese",
                 subtitle = "What role does income play in determining obesity?") +
            coord_sf(xlim = c(-67, -74), ylim = c(41, 47.4)) +
            #change background to white
            theme(panel.background = element_blank(),
           axis.line = element_line(colour = "white"),
           #changes facet background and outline
             strip.background = element_rect(colour="white", fill="white"),
           #centers and bolds title and subtitle
           plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle =element_text(hjust = 0.5),
            axis.text.x = element_blank(),
            axis.text.y = element_blank(),
            axis.ticks = element_blank(),
          legend.title = element_text(face = "bold"),
           legend.direction = "horizontal",
            legend.position = "bottom") +
          scale_fill_distiller(palette = "Oranges", direction = 1)
  
  
  #scale_fill_gradientn(colours = topo.colors(4))
      #  brewer.pal(4, “Set3”)

```

Obesity is more prevalent among men in the middle income distribution than in the lowest and highest brackets. Women on the other hand have a higher rate of obesity in the lowest income bracket and have lower rates in the top and middle distributions. While the states that make up New England have obesity rates lower than the national average, obesity in New England follows the national trend of varying with income. 




#Exercise 

As an attempt to further explain the factors which may contribute to low-income people’s high levels of obesity, we created the visualization above which displays New England’s major commuting zones, colored as the fraction of exercise each survey participant partook in during the past 30 days and sorted by each income quartile.  Since there is an association between sedentariness, poor health, and obesity, we felt that exercise rates might explain low-income obesity in New England.  

```{r,echo=FALSE, warning=FALSE, message=FALSE, fig.height= 10, fig.width= 14}
#Graph 3: Exercise by Income Quartile

#exercise rates by income
income_and_exercise_rates_NE <- income_and_health_by_cz_NE %>% 
  select(cty, statename.x, geometry, exercise_any_q1, exercise_any_q2, exercise_any_q3, exercise_any_q4) %>%
  #gathering
  gather(key = income_quartile, value = perc_exercise,
         exercise_any_q1, exercise_any_q2, exercise_any_q3, exercise_any_q4)

#labels for facets
levels = c('exercise_any_q1','exercise_any_q2','exercise_any_q3','exercise_any_q4') 
income_and_exercise_rates_NE2 <- c(exercise_any_q1 = "Quartile 1", exercise_any_q2 = "Quartile 2", exercise_any_q3 = "Quartile 3", exercise_any_q4 = "Quartile 4")

#graph average frequency of exercise rates by income
ggplot() + geom_sf(data = income_and_exercise_rates_NE, aes(fill = perc_exercise))+ 
  facet_wrap(~income_quartile, labeller = labeller(income_quartile = income_and_exercise_rates_NE2), nrow = 1) +
  labs(title = "Frequency of Exercise Per Income Quartile in New England",  fill = "Frequency of\nExercise",
       subtitle ="How does income influence one's fitness?") +
  coord_sf(xlim = c(-67, -74), ylim = c(41, 47.4)) +
  #change background to white
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "white"),
        #changes facet background and outline
        strip.background = element_rect(colour="white", fill="white"),
        #centers and bolds title and subtitle
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle =element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_text(face = "bold"),
       legend.direction = "horizontal",
       legend.position = "bottom") +
  scale_fill_distiller(palette = "YlOrBr", direction = 1)
       
  
  
  #scale_fill_gradientn(colours = rainbow(4))
  
  #cm.colors(5)
  #scale_fill_distiller(palette = "Greens")
```

Given the prior visualization results, the fact that this visualization also shows that low-exercise rates is correlated with low-income is unsurprising.  However, the reason for it is shocking.  A survey done by the health promotion group, Shape Up America suggests that it isn't a lack of knowledge that keeps poor people from exercising more, rather it is a lack of means.  Low-income people face major obstacles to exercising such not feeling safe exercising outdoors in their neighborhoods due to violence, not having access to public facilities, and the inability to afford private facilities.  This suggests, improving the poor’s access to safe exercise facilities could prove to largely effective at combating prevailing health deficits. 

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3198075/
http://articles.latimes.com/1995-11-26/news/mn-7235_1_poor-people

#Concluding Thoughts

Our findings were consistent with the sources we referenced: health inequality is related to income inequality, and low income areas tend to be less healthy than higher income areas. This trend holds true over each income quintile; in other words, the higher the income quintile, the healthier the individuals in that area. Wealthier areas in New England tended to have lower rates of obesity, lower tobacco use, more frequent exercise rates, and higher life expectancies. Interestingly, we found that women tend to live longer than men in their income quintile for all income quintiles. Even though New England is comprised of healthy states relative to states in other regions, health inequality related to income inequality is still prevalent. 


# References and Citations

* [Why are 72% of Smokers from Lower-income Communities?](https://truthinitiative.org/news/why-are-72-percent-smokers-lower-income-communities){target="_blank"}, Truth Initiative, 2018. 

* [Tobacco is a Social Justice Issue: Low-income Communities](https://truthinitiative.org/news/smoking-and-low-income-communities){target="_blank"}, Truth Initiative, 2017.  

* [Tobacco Use Among the Homeless Population: FAQ](https://publichealthlawcenter.org/sites/default/files/resources/tclc-homeless-tobacco-FAQ-2016.pdf){target="_blank"}, Public HEalth Law Center, 2016.

* [Poverty and Obesity in the U.S.](http://diabetes.diabetesjournals.org/content/60/11/2667){target="_blank"}, American Diabetes Association, 2011.

* [Adult Obesity in the United States](https://stateofobesity.org/adult-obesity/){target="_blank"}, The State of Obesity, 2018.

* [State-Specific Prevalence of Current Cigarette Smoking and Smokeless Tobacco Use Among Adults — United States, 2014](https://www.cdc.gov/mmwr/volumes/65/wr/mm6539a1.htm){target="_blank"}, Centers for Disease Control and Prevention, 2017.

* [Poor Lack Means to Exercise, Survey Says : Health: The less people earn, the more likely they are to say that their neighborhood is unsafe for workouts, poll finds.](http://articles.latimes.com/1995-11-26/news/mn-7235_1_poor-people){target="_blank"}, Los Angeles Times, 1995. 