---
title: "The Effects of Smoking, Obesity, and Exercise on Life Expectancy"
author: "Group 14: Sarah Albert, Morgan Barney, and  Ana Porro"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: kable
---

<!--
Comments in HTML are like this! 
-->

```{r, include=FALSE, warning=FALSE, message=FALSE}
#Load Packages
library(ggplot2)
library(tidyverse)
library(leaflet)
library(sf)
library(USAboundaries)
library(maps)
library(dplyr)
library(plyr)
library(readr)
library(knitr)
library(htmltools)
library(htmlwidgets)
library(RColorBrewer)
library(magrittr)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Data Wrangling

#Load in shape file
cz_1990 <- "cz1990_shapefile" %>%
  read_sf()

#load in data sets
health_ineq_online_table_11 <- read_csv("health_ineq_online_table_11.csv")

health_ineq_online_table_12 <- read_csv("health_ineq_online_table_12.csv")

#select important variables from health_ineq_online_table_11
#this is so the full join goes nicely
health_ineq_online_table_11_v2 <- select(health_ineq_online_table_11, cty, statename, le_raceadj_q1_F, le_agg_q1_F, 
                                         le_raceadj_q2_F, le_agg_q2_F, le_raceadj_q3_F, le_agg_q3_F, 
                                         le_raceadj_q4_F, le_agg_q4_F, le_raceadj_q1_M, le_agg_q1_M, 
                                         le_raceadj_q2_M, le_agg_q2_M, le_raceadj_q3_M, le_agg_q3_M, 
                                         le_raceadj_q4_M, le_agg_q4_M)

#combine health_ineq_online_table_11 and health_ineq_online_table_12 by county (cty) and view
income_and_health_by_cty <- full_join(health_ineq_online_table_11_v2, health_ineq_online_table_12, by = "cty")

#combine income_and_health_by_cty and shapefile by communting zone (cz)
income_and_health_by_cz <- full_join(income_and_health_by_cty, cz_1990, by = "cz")

#filtering for NE States
income_and_health_by_cz_NE <- income_and_health_by_cz %>% filter(statename.x == "Massachusetts" | 
                                                                   statename.x == "Maine" |
                                                                   statename.x == "New Hampshire" |
                                                                   statename.x == "Vermont" |
                                                                   statename.x == "Rhode Island")
#View(income_and_health_by_cz_NE)
```




# Introduction

To further investigate the possible factors which may influence low-income peoples’ relatively depressed life expectancies, we used survey data from the Behavioral Risk Factor Surveillance System (BRFSS) during the year 2000, found in Chetty et. al’s data set.^[Chetty, Raj, et al. “The Association between Income and Life Expectancy in the United States, 2001 - 2014.” The Equality of Opportunity Project, Opportunity Insights, 2016, www.equality-of-opportunity.org/data/index.html#health.]The BRFSS is the nation's premier provider of health-related telephone surveys which collect state data about U.S. residents regarding their health-related behaviors.^[“Behavioral Risk Factor Surveillance System.” Centers for Disease Control and Prevention, National Center for Chronic Disease Prevention and Health Promotion , Division of Population Health, 16 May 2014, www.cdc.gov/brfss/index.html"] The three health indicators we considered by income quartile were tobacco usage, obesity rates, and exercise rates.




# Tobacco

A 2016 report on the economics of tobacco published by the National Cancer Institute and World Health Organization found that a significant amount of health disparities between the rich and poor can be attributed to the differences in tobacco habits between income groups.^[“Monograph 21: The Economics of Tobacco and Tobacco Control.” National Cancer Institute, Division of Cancer Control & Population Sciences, The National Cancer Institute and the World Health Organization, 2016, cancercontrol.cancer.gov/brp/tcrb/monographs/21/index.html.]
	
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height= 10/2, fig.width= 16/2}
#Graph 4: Smokers by Income Quartile

income_and_smoking_rates_NE <- income_and_health_by_cz_NE %>% 
  select(cty, statename.x, geometry, le_agg_q1_F, 
         cur_smoke_q1, cur_smoke_q2, cur_smoke_q3, cur_smoke_q4) %>%
  #gathering
  gather(key = income_quartile, value = perc_smokers,
         cur_smoke_q1, cur_smoke_q2, cur_smoke_q3, cur_smoke_q4)

#labels for smoking facets
smoke_quartile_labels <- c(cur_smoke_q1 = "Quartile 1", cur_smoke_q2 = "Quartile 2", cur_smoke_q3 = "Quartile 3", cur_smoke_q4 = "Quartile 4") 

#graph smoking rates by income

ggplot() + geom_sf(data = income_and_smoking_rates_NE, aes(fill = perc_smokers))+ 
  facet_wrap(~income_quartile, labeller = labeller(income_quartile = smoke_quartile_labels), nrow = 1) +
  coord_sf(xlim = c(-67, -74), ylim = c(41, 47.4)) +
  #changes color palette
  #scale_fill_brewer(palette="YlGn") +
  #change background to white
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "white"),
        #changes facet background and outline
        strip.background = element_rect(colour="white", fill="white"),
        #centers and bolds title and subtitle
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle =element_text(hjust = 0.5),
        #gets rid of x and y axis labels
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_text(face = "bold"),
        legend.direction = "horizontal",
       legend.position = "bottom",
       legend.text = element_text(size = rel(.5)),
       strip.text = element_text(face = "bold")) +
  #labels
  labs(title= "Smoking Rates Per Income Quartile in New England", fill = "Share of\nSmokers", subtitle ="How does income influence smoking?") +
   scale_fill_distiller(palette = "Reds", direction = 1)
       
  
  
  #scale_fill_gradientn(colours = heat.colors(4))
  
```

The data for New England states embodies this trend. Smoking rates for the lowest quartile are much higher than those in the highest quartile. This could be a result of historic targeted marketing of tobacco products to low-income individuals by attaching coupons to food stamps and giving out free cigarettes in housing projects.^[“Why Are 72% of Smokers from Lower-Income Communities?” Truth Initiative: Inspiring  Tobacco Free Lives, Truth Initiative, 24 Jan. 2018, truthinitiative.org/news/why-are-72-percent-smokers-lower-income-communities] Additionally, the tobacco industry fights to keep prices low so that low-income people can continue to foster their nicotine addiction by spending billions of dollars each year to fight cigarette taxes and price increases.^[Truth Initiative, “Why Are 72% of Smokers from Lower-Income Communities?”] Further, low-income communities have less access to resources to help them quit smoking, and are exposed to less education on the health hazards of smoking.^[NCI and WHO,"The Economics of Tobacco and Tobacco Control"]

# Obesity

The U.S. is one of the wealthiest countries in the world yet it has one of the highest rates of obesity. The statistics are shocking: one-third of the American population is obese, while another third is overweight^[Levine, James A. “Poverty and Obesity in the U.S.” American Diabetes Association: Diabetes Journals, American Diabetes Association, Nov. 2011, diabetes.diabetesjournals.org/content/60/11/2667.] Given the negative health effects and shortened life expectancy associated with obesity, we investigated whether obesity rates varied across income quartiles similar to life expectancy. The graphic below examines the prevalence of obesity per income quartile in New England.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height= 10/2, fig.width= 16/2}
#Graph: Obesity by Income Quartile

#obesity rates by income
income_and_obesity_rates_NE <- income_and_health_by_cz_NE %>% 
  select(cty, statename.x, geometry, le_agg_q1_F, 
         bmi_obese_q1, bmi_obese_q2, bmi_obese_q3, bmi_obese_q4) %>%
  #gathering
  gather(key = income_quartile, value = perc_obese,
         bmi_obese_q1, bmi_obese_q2, bmi_obese_q3, bmi_obese_q4)

#code as factor to order facets
income_and_obesity_rates_NE$income_quartile_GG = factor(income_and_obesity_rates_NE$income_quartile, levels=c('bmi_obese_q1','bmi_obese_q2','bmi_obese_q3','bmi_obese_q4'))
                                                        #lables for facets
                                                        levels=c('bmi_obese_q1','bmi_obese_q2','bmi_obese_q3','bmi_obese_q4') 
                                                        
income_and_obesity_rates_NE2 <- c(bmi_obese_q1 = "Quartile 1", bmi_obese_q2 = "Quartile 2", bmi_obese_q3 = "Quartile 3", bmi_obese_q4 = "Quartile 4")

#factor bmi_obesity for color scale

#graph obsesity rates
ggplot() + geom_sf(data = income_and_obesity_rates_NE, aes(fill = perc_obese))+ 
            facet_wrap(~income_quartile, labeller = labeller(income_quartile = income_and_obesity_rates_NE2), nrow = 1) +
            labs(title = "Obesity Per Income Quartile in New England",  fill = "Percent Obese",
                 subtitle = "What role does income play in determining obesity?") +
            coord_sf(xlim = c(-67, -74), ylim = c(41, 47.4)) +
            #change background to white
            theme(panel.background = element_blank(),
           axis.line = element_line(colour = "white"),
           #changes facet background and outline
             strip.background = element_rect(colour="white", fill="white"),
           #centers and bolds title and subtitle
           plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle =element_text(hjust = 0.5),
            axis.text.x = element_blank(),
            axis.text.y = element_blank(),
            axis.ticks = element_blank(),
          legend.title = element_text(face = "bold"),
           legend.direction = "horizontal",
            legend.position = "bottom",
           legend.text = element_text(size = rel(.5)),
          strip.text = element_text(face = "bold")) +
          scale_fill_distiller(palette = "Oranges", direction = 1)

```

The data above shows that obesity rates decrease as income increases in the New England area. Therefore, this analysis points to a relationship between obesity and poverty. The Mayo Clinic proposes a reason behind this trend: often, poverty-stricken areas have such low access to healthy foods that they are called "food deserts".^[Levine,“Poverty and Obesity in the U.S.”] Since there is an association between sedentariness, poor health, and obesity, we felt that exercise rates might explain low-income obesity in New England. We investigate this in the next section.


# Exercise 

As an attempt to further explain the factors which may contribute to low-income people’s high levels of obesity, we created the visualization below which displays New England’s major commuting zones, colored as a fraction of the exercise each person surveyed participated in during the past 30 days, grouped by income quartile. 

```{r,echo=FALSE, warning=FALSE, message=FALSE, fig.height= 10/2, fig.width= 16/2}
#Graph 3: Exercise by Income Quartile

#exercise rates by income
income_and_exercise_rates_NE <- income_and_health_by_cz_NE %>% 
  select(cty, statename.x, geometry, exercise_any_q1, exercise_any_q2, exercise_any_q3, exercise_any_q4) %>%
  #gathering
  gather(key = income_quartile, value = perc_exercise,
         exercise_any_q1, exercise_any_q2, exercise_any_q3, exercise_any_q4)

#labels for facets
levels = c('exercise_any_q1','exercise_any_q2','exercise_any_q3','exercise_any_q4') 
income_and_exercise_rates_NE2 <- c(exercise_any_q1 = "Quartile 1", exercise_any_q2 = "Quartile 2", exercise_any_q3 = "Quartile 3", exercise_any_q4 = "Quartile 4")

#graph average frequency of exercise rates by income
ggplot() + geom_sf(data = income_and_exercise_rates_NE, aes(fill = perc_exercise))+ 
  facet_wrap(~income_quartile, labeller = labeller(income_quartile = income_and_exercise_rates_NE2), nrow = 1) +
  labs(title = "Frequency of Exercise Per Income Quartile in New England",  fill = "Frequency of\nExercise",
       subtitle ="How does income influence one's fitness?") +
  coord_sf(xlim = c(-67, -74), ylim = c(41, 47.4)) +
  #change background to white
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "white"),
        #changes facet background and outline
        strip.background = element_rect(colour="white", fill="white"),
        #centers and bolds title and subtitle
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle =element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_text(face = "bold"),
       legend.direction = "horizontal",
       legend.position = "bottom",
       strip.text = element_text(face = "bold"),
      legend.text = element_text(size = rel(.5))) +
  scale_fill_distiller(palette = "YlOrBr", direction = 1)
       
  
  #scale_fill_gradientn(colours = rainbow(4))
  
  #cm.colors(5)
  #scale_fill_distiller(palette = "Greens")
```

Given the prior visualization results, the fact that this graphic also shows that low-exercise rates are correlated with low-income is not surprising. However, the reason behind it is shocking. A survey conducted by the health promotion group, Shape Up America suggests that what keeps poor people from exercising more is not a lack of knowledge, but a lack of means.^[Dreyfuss, Ira. “Poor Lack Means to Exercise, Survey Says : Health: The Less People Earn, the More Likely They Are to Say That Their Neighborhood Is Unsafe for Workouts, Poll Finds.” Los Angeles Times, Los Angeles Times, 26 Nov. 1995, articles.latimes.com/1995-11-26/news/mn-7235_1_poor-people.] Low-income people face major obstacles to exercising such as not feeling safe exercising outdoors in their neighborhoods because of violence, not having access to public facilities, and the inability to afford private facilities.^[Dreyfus, "Poor Lack Means to Exercise..."] This suggests, improving the poor’s access to safe exercise facilities could prove largely effective at combating prevailing health deficits. 


# Concluding Thoughts

Our findings were consistent with national trends in scholarship: health inequality is related to income inequality, and low income areas tend to be less healthy than higher income areas. Wealthier areas in New England tended to have higher life expectancies, lower tobacco use, lower rates of obesity, and higher exercise rates. While universal healthcare might be the solution to addressing these national problems, this analysis suggests that introducing targeted solutions such as offering affordable nicotine-quitting resources, access to healthier foods, and safe public health facilities, may prove more effective at addressing health impacting behaviors.


# References and Citations


* [“Adult Obesity in the United States.” The State of Obesity, Robert Wood Johnson Foundation, Sept. 2018, stateofobesity.org/adult-obesity/.](https://stateofobesity.org/adult-obesity/){target="_blank"}

* [“Behavioral Risk Factor Surveillance System.” Centers for Disease Control and Prevention, National Center for Chronic Disease Prevention and Health Promotion , Division of Population Health, 16 May 2014.](www.cdc.gov/brfss/index.html){target="_blank"}

* [Chetty, Raj, et al. “The Association between Income and Life Expectancy in the United States, 2001 - 2014.” The Equality of Opportunity Project, Opportunity Insights, 2016](www.equality-of-opportunity.org/data/index.html#health){target="_blank"}

* [Chetty, Raj, et al. “The Association between Income and Life Expectancy in the United States, 2001 - 2014.” Jama Network, American Medical Association, 2016](https://jamanetwork.com/journals/jama/fullarticle/2513561?guestAccessKey=4023ce75-d0fb-44de-bb6c-8a10a30a6173){target="_blank"}


* [Dreyfuss, Ira. “Poor Lack Means to Exercise, Survey Says : Health: The Less People Earn, the More Likely They Are to Say That Their Neighborhood Is Unsafe for Workouts, Poll Finds.” Los Angeles Times, Los Angeles Times, 26 Nov. 1995](http://articles.latimes.com/1995-11-26/news/mn-7235_1_poor-people){target="_blank"}

* [“Monograph 21: The Economics of Tobacco and Tobacco Control.” National Cancer Institute, Division of Cancer Control & Population Sciences, The National Cancer Institute and the World Health Organization, 2016](cancercontrol.cancer.gov/brp/tcrb/monographs/21/index.html){target="_blank"} 

* [Levine, James A. “Poverty and Obesity in the U.S.” American Diabetes Association: Diabetes Journals, American Diabetes Association, Nov. 2011.](diabetes.diabetesjournals.org/content/60/11/2667.){target="_blank"} 

* [“Tobacco and Social Justice: Smoking in Low-Income Communities.” Truth Initiative: Inspiring Tobacco Free Lives, Truth Initiative, 31 Jan. 2017](https://truthinitiative.org/news/smoking-and-low-income-communities){target="_blank"}

* [“Tobacco Use Among the Homeless Population: FAQ.” Public Health Law Center, Public Health Law Center, Dec. 2016](publichealthlawcenter.org/sites/default/files/resources/tclc-homeless-tobacco-FAQ-2016.pdf){target="_blank"}

* [“Why Are 72% of Smokers from Lower-Income Communities?” Truth Initiative: Inspiring Tobacco Free Lives, Truth Initiative, 24 Jan. 2018, truthinitiative.org/news/why-are-72-percent-smokers-lower-income-communities.](truthinitiative.org/news/why-are-72-percent-smokers-lower-income-communities){target="_blank"}

